Untitled

### 2024-03-06 models_appx.Rmd
random updates that I don't remember

## PRELIMINARY MODELS

IMPORTANCE OF URBAN? ADD EXTRA PLOT HERE? BLK / URBAN?

1) jim crow salience (can we define as brutal coupling of formal / informal? less state capacity built for maintenance / strategic underinvestment)
2) exclusion / formal soc control salience in urban non-S

employment trajectories look kinda log-linear. almost certainly non-linear, so need a curve on trajectories regardless. log probably solves it well enough.

q1: salience of settler expropriotion -> state formation

four models, each employment outcome + children in poverty institutions

start with lm for log(rates)

trouble fitting models for sparse data (e.g. social workers, children in gq), but works fine and in expected directions for teachers, police.

switch to brms for estimation on tricky ones

```{r}
### FIT PROBLEM WAS WITH EARLY YEARS NEAR ZERO AND LOW VARIANCE
model_dat <- dat %>%
  filter(year>=1880) %>%
  mutate(police_rate = (police + 1) / tot_pop * 1e5,
         socwork_rate = (socwork + 1) / tot_pop * 1e5,
         teacher_rate = (teachers + 1) / tot_pop * 1e5,
         child_inst_rate = (child_inst_poor_disabled + 0.1) /
           children * 1e5,
         prop_blk = (blk + 1) / tot_pop,
         prop_urban_immig = urban_immig_pop / tot_pop,
         hsclaim_prop = hs_claims / tot_pop,
         year = scale(year))

m_police<-brm(log(police_rate) ~
                 scale(hsclaim_prop) +
                 scale(prop_urban_immig) +
                 scale(prop_blk) +
                 scale(log(blk_sharecrop + 1)) +
                 year +
                 (1 + year|state),
               data = model_dat)

m_teacher<-brm(log(teacher_rate) ~
                 scale(hsclaim_prop) +
                 scale(prop_urban_immig) +
                 scale(prop_blk) +
                 scale(log(blk_sharecrop + 1)) +
                 year +
                 (1 + year|state),
               data = model_dat)

m_socwork<-brm(log(socwork_rate) ~
                 scale(hsclaim_prop) +
                 scale(prop_urban_immig) +
                 scale(prop_blk) +
                 scale(log(blk_sharecrop + 1)) +
                 year +
                 (1 + year|state),
               data = model_dat)

sc_teacher<-lmer(log(teacher_rate) ~
              scale(hsclaim_prop) +
              year +
              (1 + year|state),
            data = model_dat,
            control=lmerControl(optimizer="nloptwrap",
                                optCtrl=list(maxfun=2e5)))

sc_socwork<-brm(log(socwork_rate) ~
              scale(hsclaim_prop) +
              year +
              (1 + year|state),
            data = model_dat)

sc_gq<-brm(log(child_inst_rate) ~
              scale(hsclaim_prop) +
              year +
              (1 + year|state),
            data = model_dat)

jc_police<-lmer(log(police_rate) ~
              scale(prop_blk) +
              year +
              (1 + year|state),
            data = model_dat,
            control=lmerControl(optimizer="nloptwrap",
                                optCtrl=list(maxfun=2e5)))

jc_teacher<-lmer(log(teacher_rate) ~
              scale(prop_blk) +
              year +
              (1 + year|state),
            data = model_dat,
            control=lmerControl(optimizer="nloptwrap",
                                optCtrl=list(maxfun=2e5)))

jc_socwork<-lmer(log(socwork_rate) ~
              scale(prop_blk) +
              year +
              (1 + year|state),
            data = model_dat,
            control=lmerControl(optimizer="nloptwrap",
                                optCtrl=list(maxfun=2e5)))

jc_gq<-lmer(log(child_inst_rate) ~
              scale(prop_blk) +
              year +
              (1 + year|state),
            data = model_dat,
            control=lmerControl(optimizer="nloptwrap",
                                optCtrl=list(maxfun=2e5)))
```

This all looks fine. How do I want to visualize this?

we'll eventually do trajectory visuals for each regime type - what is the expected trajectory under different regimes, where do we end up in 1930? Think about how to handle to intercepts / slopes. What would a zero mean? pulling out everything not directly in national trend and regime measures. is that what I'm after? I guess I really do just want the regime-specific trend lines, in which case E[rate | regime] should zero out the individual idioscyncracies. it's oversimplified, but that should be ok. Just be super clear on what the iterpretation is on prediction. it's not a hypothetical state, its the mean expectation for the slope of the regime var.

model is

\[\hat{Y} = \alpha + \beta \textrm{Regime} + \delta year + \epsilon\]

Here we are very much interested just in $\beta$, the rest are nuissance parameters. that works. can vis e[state type] under diffferent REGIME sets.

Q: are there places with competing regime sets? CA, WA, others? where we need a multivariate regime model?

```{r}
new_dat<-data.frame(year =  c(1.3, 1.3, 1.3),
                    hsclaim_prop = c(0, 0.05, 0.15))
t<-predict(sc_police,
           newdata = new_dat,
           re.form = ~0,
           type = "link")

t1<-predict(sc_teacher,
           newdata = new_dat,
           re.form = ~0,
           type = "link")

exp(t1)

>>>>

explore model fit under alternate specifications. look at bivars and functional relationships b/w outcomes and focal predictors

describe prior selection

```{r}
## Diagnostics for model fit
# focus on zero inflated processes
# and parsimony / convergence trade offs
# using default brms priors for ease
# log(children) not improving teacher fit. leaving it out

mdat<-pop %>% 
  left_join(dist) %>% 
  left_join(bs)

###### teachers - easiest outcome given power
### baseline model as reported in eq (1) in ms
t_m1<-brm(
  teachers ~ 
    scale(I(blk_sharecrop / tot_pop)) +
    scale(I(urban_immig_pop / tot_pop)) +
    scale(morill_acres_cum + homestead_acres_cum) +
    factor(year) + 
    (1|STUSPS) + 
    offset(log(tot_pop)),
  family = negbinomial(),
  data = mdat
)

### prior visual
# teachers_m1 %>% 
#   prior_summary() %>% 
#   parse_dist(prior) %>% 
#   ggplot(aes(y = class, dist = .dist, args = .args)) + 
#   stat_dist_halfeye()

### add regime zero inflation / threshold adjustments
t_m2<-brm(
  teachers ~ 
    scale(I(blk_sharecrop / tot_pop)) +
    I(blk / tot_pop > 0.3) +  
    scale(I(urban_immig_pop / tot_pop)) +
    I(urban_pop>2e5) + 
    scale(morill_acres_cum + homestead_acres_cum) +
    I((morill_acres_cum + homestead_acres_cum)>0) + 
    log(children) + 
    factor(year) + 
    (1|STUSPS) + 
    offset(log(tot_pop)),
  family = negbinomial(),
  data = mdat
)

t_m3<-brm(
  teachers ~ 
    scale(I(blk_sharecrop / tot_pop)) +
    scale(I(urban_immig_pop / tot_pop)) +
    scale(morill_acres_cum + homestead_acres_cum) +
    log(children) + 
    factor(year) + 
    (1|STUSPS) + 
    offset(log(tot_pop)),
  family = negbinomial(),
  data = mdat
)

loo_compare(loo(t_m1), loo(t_m2))

#### cops

p_m1<-brm(
  police ~ 
    scale(I(blk_sharecrop / tot_pop)) +
    scale(I(urban_immig_pop / tot_pop)) +
    scale(morill_acres_cum + homestead_acres_cum) +
    log(children) + 
    factor(year) + 
    (1|STUSPS) + 
    offset(log(tot_pop)),
  family = negbinomial(),
  data = mdat
)

### add regime zero inflation / threshold adjustments
p_m2<-brm(
  police ~ 
    scale(I(blk_sharecrop / tot_pop)) +
    I(blk / tot_pop > 0.3) +  
    scale(I(urban_immig_pop / tot_pop)) +
    I(urban_pop>2e5) + 
    scale(morill_acres_cum + homestead_acres_cum) +
    I((morill_acres_cum + homestead_acres_cum)>0) + 
    log(children) + 
    factor(year) + 
    (1|STUSPS) + 
    offset(log(tot_pop)),
  family = negbinomial(),
  data = mdat
)

loo_compare(loo(p_m1), loo(p_m2))

#### loo shows no meaningful difference bw m1, m2 for both teachers and cops

###### GQ MODELS

mdat <- mdat %>% 
  mutate(inst_gq = inst_mental + 
           inst_poor_disabled)

prison_m1<-brm(
  floor(inst_correctional) ~ 
    scale(I(blk_sharecrop / tot_pop)) +
    scale(I(urban_immig_pop / tot_pop)) +
    scale(morill_acres_cum + homestead_acres_cum) +
    log(children) + 
    factor(year) + 
    (1|STUSPS) + 
    offset(log(tot_pop)),
  family = negbinomial(),
  data = mdat %>% 
    filter(year!=1910))

prison_m2<-brm(
  floor(inst_correctional) ~ 
    scale(I(blk_sharecrop / tot_pop)) +
    I(blk / tot_pop > 0.3) +  
    scale(I(urban_immig_pop / tot_pop)) +
    I(urban_pop>2e5) + 
    scale(morill_acres_cum + homestead_acres_cum) +
    I((morill_acres_cum + homestead_acres_cum)>0) + 
    log(children) + 
    factor(year) + 
    (1|STUSPS) + 
    offset(log(tot_pop)),
  family = negbinomial(),
  data = mdat %>% 
    filter(year!=1910))

loo_compare(loo(prison_m1), loo(prison_m2))

### no meaningful difference in models according to loo
### NJ 1900 shows NA
### looks like failed joins. investigate!
### ballparking numbers for prisoners against https://bjs.ojp.gov/content/pub/pdf/p2581.pdf
### not too bad! 1925 n is 91,689; 1930 is 129,453. I report 215682 for 1930, 79073 for 1920 
# year gqtype  gq_pop
# 1  1880      2  49272 
# 2  1900      2  71491 
# 3  1910      2 135323.
# 4  1920      2  79073 
# 5  1930      2 215682 
### 1910 is discontinuous - that's the sample year, not full census. check docs again
# From IPUMS
# User Note: The Census Bureau did not collect information on the type of group quarters in the 1910 census. 
# This is reflected in the 1910 complete count file, where all group-quarters units have been assigned a GQ code of 4 (other group quarters). 
# Values for GQ, GQTYPE, and GQFUNDS in the 1910 1% and 1.4% samples have been imputed.

welfgq_m1<-brm(
  floor(inst_gq) ~ 
    scale(I(blk_sharecrop / tot_pop)) +
    scale(I(urban_immig_pop / tot_pop)) +
    scale(morill_acres_cum + homestead_acres_cum) +
    log(children) + 
    factor(year) + 
    (1|STUSPS) + 
    offset(log(tot_pop)),
  family = negbinomial(),
  data = mdat %>% 
    filter(year!=1910))

welfgq_m2<-brm(
  floor(inst_gq) ~ 
    scale(I(blk_sharecrop / tot_pop)) +
    I(blk / tot_pop > 0.3) +  
    scale(I(urban_immig_pop / tot_pop)) +
    I(urban_pop>2e5) + 
    scale(morill_acres_cum + homestead_acres_cum) +
    I((morill_acres_cum + homestead_acres_cum)>0) + 
    log(children) + 
    factor(year) + 
    (1|STUSPS) + 
    offset(log(tot_pop)),
  family = negbinomial(),
  data = mdat %>% 
    filter(year!=1910))

loo_compare(loo(welfgq_m1), loo(welfgq_m2))

# boarding schools models use 1890 - 1920 time period
# models need to be specified differntly because of census mismatch
# may need to increase samples to up ESS
# but for a simple descriptive, this looks fine IMO

bs_mdat<-dist %>% 
  left_join(bs) %>% 
  left_join(agents %>% 
              filter(year!=1910)) %>% # 1910 is incomplete data
  filter(year>=1900, year<=1930) # 1900 - 1920 for enrollment / attendance, 1900 - 1930 for school count
#### maybe run this zero inflated,

######## BIVARIATES AND DIAGNOSTICS
ggplot(bs_mdat %>% 
         filter(boarding_schools>0),
       aes(x = log(homestead_acres_cum + morill_acres_cum),
           y = boarding_schools,
           color = STUSPS)) + 
  geom_point() + 
  facet_wrap(~division)

### stability in boarding_schools - increase in acres distributed, not much to model other than average
### only action is decrease in time series at end, so that's a bit funky

ggplot(bs_mdat %>% 
         filter(boarding_schools>0,
                morill_acres_cum>0),
       aes(x = log(homestead_acres_cum + morill_acres_cum), 
           y = boarding_n_acc,
           color = STUSPS)) + 
  geom_point() + 
  facet_wrap(~division)

### similar look here. some increases in bs pops, e.g. in SD, AZ, OK, but mostly stable over the period
### some clear increases and maybe log-linear things going on, but don't want to overfit.

ggplot(bs_mdat,
       aes(x = boarding_schools,
           fill = homestead_acres_cum>0))+ 
  geom_density(alpha = 0.2)

#### OK - zero inflation is way to go, but there's no 1900 - 1930 variation in the binary
### boarding_schools>0 within states, so panel model makes no sense for binary outcome
### subset to start of period condition, then conditional on nonzero in 1900, look at count

#### LEANING TOWARD THE SCHOOL COUNT ONLY APPROACH. THE CAPACITY / ENROLLMENT NUMBERS NEED MORE WORK I THINK
### START HERE TOMORROW. FOLD THE REVISED NO LOG(KIDS) MODEL IN, FOLD THE ZINFL MODEL IN

bs_m1_z<-brm(boarding_schools>0 ~ 
               scale(I(blk_sharecrop / tot_pop)) +
               scale(I(urban_immig_pop / tot_pop)) +
               scale(morill_acres_cum + homestead_acres_cum),
             family = bernoulli(link = "logit"),
             data = mdat %>% 
               filter(year == 1900))

bs_m1<-brm(boarding_schools ~ 
             scale(I(blk_sharecrop / tot_pop)) +
             scale(I(urban_immig_pop / tot_pop)) +
             scale(morill_acres_cum + homestead_acres_cum) +
             factor(year) +
             (1|STUSPS),
           family = negbinomial(),
           data = mdat %>% 
             filter(boarding_schools>0))

### zero infl model for capacity

cap_m1_z<-brm(boarding_n_acc>0 ~ 
            scale(I(blk_sharecrop / tot_pop)) +
            scale(I(urban_immig_pop / tot_pop)) +
            scale(morill_acres_cum + homestead_acres_cum) ,
            # factor(year) + 
            # (1|STUSPS),
          family = bernoulli(link = "logit"),
          data = mdat %>% 
            filter(year == 1900))

cap_m1<-brm(boarding_n_acc ~ 
            scale(I(blk_sharecrop / tot_pop)) +
            scale(I(urban_immig_pop / tot_pop)) +
            scale(morill_acres_cum + homestead_acres_cum) +
            factor(year) +
            (1|STUSPS),
          family = negbinomial(),
          data = mdat %>% 
            filter(boarding_n_acc>0))

#### model vis
# bivar - looks like it's just a binary relationship, the linear component isn't adding much here
ggplot(bs_mdat %>% 
         filter(boarding_n_acc >0),
       aes(x = morill_acres_cum + homestead_acres_cum,
           y = boarding_n_acc,
           color = STUSPS)) + 
  geom_point()

ggplot(bs_mdat %>% 
         filter(boarding_n_acc >0),
       aes(x = morill_acres_cum + homestead_acres_cum,
           y = boarding_schools,
           color = STUSPS)) + 
  geom_point()
